---
title: 'Lab 2: Diagnostic Tests'
author: "M. Rosário Oliveira & Catarina Loureiro - Biostatistics (LMAC/MECD/MMAC)"
date: "`r format(Sys.time(), '%B %Y')`"
output:
  html_document:
    toc: true
    df_print: paged
  pdf_document:
    toc: true
    number_sections: true
    fig_width: 7
    fig_height: 6
    fig_caption: true
    extra_dependencies:
      caption: labelfont={bf}
      hyperref: null
      xcolor: null
---

------------------------------------------------------------------------

Report your answers to the following questions in an R Markdown file. Submit your answers by the 20th of March 2025, 23:59, on the Fenix webpage.

Add your group number and the students' name and number of its members in the file:

-   **Group Number:**

-   **Name and Number - Group Member 1: Mariana Henriques 103165**

-   **Name and Number - Group Member 2:**

------------------------------------------------------------------------

Consider the following dataset referring to the results of 3 diagnostic tests.

| Test A   | Test B   | Test C   | Frequency observed |
|----------|----------|----------|--------------------|
| Positive | Positive | Positive | 34                 |
| Positive | Positive | Negative | 12                 |
| Positive | Negative | Positive | 19                 |
| Negative | Positive | Positive | 11                 |
| Positive | Negative | Negative | 23                 |
| Negative | Positive | Negative | 59                 |
| Negative | Negative | Positive | 21                 |
| Negative | Negative | Negative | 1513               |

### Exercise 1: Measures of performance

Assuming that test A is an imperfect gold standard, estimate the following performance measures of Test C:

1.  Sensitivity;
2.  Specificity;
3.  Positive predictive value;
4.  Negative predicted value;
5.  Youden's index;
6.  Positive Diagnostic Likelihood Ratio;
7.  Negative Diagnostic Likelihood Ratio;
8.  Prevalence.

Interpret your findings.

```{r}

# Create the dataset as provided
data <- data.frame(
  TestA = c("Positive", "Positive", "Positive", "Negative", "Positive", "Negative", "Negative", "Negative"),
  TestB = c("Positive", "Positive", "Negative", "Positive", "Negative", "Positive", "Negative", "Negative"),
  TestC = c("Positive", "Negative", "Positive", "Positive", "Negative", "Negative", "Positive", "Negative"),
  Frequency = c(34, 12, 19, 11, 23, 59, 21, 1513)
)

# Considering Test A as the imperfect gold standard
# Sensitivity, Specificity, PPV, NPV, Youden's index, likelihood ratios, prevalence

# Create contingency table for Test C vs Test A
library(dplyr)

# True Positives, False Positives, True Negatives, False Negatives
TP <- sum(data %>% filter(TestA == "Positive" & TestC == "Positive") %>% pull(Frequency))
FN <- sum(data %>% filter(TestA == "Positive" & TestC == "Negative") %>% pull(Frequency))
FP <- sum(data %>% filter(TestA == "Negative" & TestC == "Positive") %>% pull(Frequency))
TN <- sum(data %>% filter(TestA == "Negative" & TestC == "Negative") %>% pull(Frequency))

# Sensitivity
Sensitivity <- TP / (TP + FN)

# Specificity
Specificity <- TN / (TN + FP)

# Positive predictive value (PPV)
PPV <- TP / (TP + FP)

# Negative predictive value (NPV)
NPV <- TN / (TN + FN)

# Youden's Index
Youdens_index <- Sensitivity + Specificity - 1

# Positive Diagnostic Likelihood Ratio (DLR+)
DLR_positive <- Sensitivity / (1 - Specificity)

# Negative Diagnostic Likelihood Ratio (DLR-)
DLR_negative <- (1 - Sensitivity) / Specificity

# Prevalence (according to Test A)
Prevalence <- (TP + FN) / sum(data$Frequency)

# Results
results <- list(
  Sensitivity = Sensitivity,
  Specificity = Specificity,
  Positive_Predictive_Value = PPV,
  Negative_Predictive_Value = NPV,
  Youdens_index = Youdens_index,
  Positive_Likelihood_Ratio = DLR_positive,
  Negative_Likelihood_Ratio = DLR_negative,
  Prevalence = Prevalence
)

# Print the results
print(results)

```

The Test C has approximately 60% of sensitivity, which means that it detects 60.2% of patients with the disease (true positives), but approximately 40% with the disease go undetected (false negatives).

The Test C has a specificity of 98%, approximately, which means it correctly identifies 98% of the people who do not have the disease, but it does not pick up the 2% who receive false positives.

The Test C has a positive predictive value of 62.4%, approximately, which means among individuals who test positive on Test C, about 63.4% truly have the disease (true positives) and about 37.6% of positives may actually not have the disease (false positives).

The Test C has a negative predictive value of 97.8%, approximately, which means among individuals who test negative on Test C, about 97.8% don't have the disease (true negatives).

The Test C has a Youden's Index of 0.58, approximately, which suggests a moderate overall diagnostic performance. The closer to 1, the better. This is, test C has good overall discrimination ability but still has room for improvement, particularly in sensitivity.

The Test C has a positive diagnostic likelihood ratio of 30.19, approximately. This means that a positive result on Test C is approximately 30 times more likely to occur in an individual with the condition than in one without it.

The Test C has a negative diagnostic likelihood ratio of 0.41,approximately. This means that a negative result on Test C slightly reduces the likelihood of the condition (0.41 times).

The Test C has a prevalence og 5.2%, approximately, which means around 5.2% of your studied population has the condition, based on your imperfect gold standard (Test A).

------------------------------------------------------------------------

### Exercise 2: Confidence intervals for binomial proportions

Compute the confidence intervals at a 95% confidence level for the sensitivity of Test C (using Test A as an imperfect reference), according to the following methods:

1.  Wald test;
2.  Clopper-Pearson;
3.  Wilson.

Visualize your findings. Get inspiration from:

<https://www.geeksforgeeks.org/how-to-plot-a-confidence-interval-in-r/>

Compare and discuss your findings.

```{r}
library(binom)
library(ggplot2)

#The function binom.confint() from the "binom" package in R calculates confidence intervals for binomial proportions, and the default alpha is 0.05, meaning it computes a 95% confidence interval

wald_ci <- binom.confint(TP, TP + FN, method = "asymptotic")
clopper_pearson_ci <- binom.confint(TP, TP + FN, method = "exact")
wilson_ci <- binom.confint(TP, TP + FN, method = "wilson")


print(paste("Sensitivity:", Sensitivity))
print("Wald CI:")
print(wald_ci)
print("Clopper-Pearson CI:")
print(clopper_pearson_ci)
print("Wilson CI:")
print(wilson_ci)

ci_data <- data.frame(
  Method = c("Wald", "Clopper-Pearson", "Wilson"),
  Lower = c(wald_ci$lower, clopper_pearson_ci$lower, wilson_ci$lower),
  Upper = c(wald_ci$upper, clopper_pearson_ci$upper, wilson_ci$upper)
)

ggplot(ci_data, aes(x = Method, y = (Lower + Upper) / 2)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = Lower, ymax = Upper), width = 0.2) +
  labs(title = "95% Confidence Intervals for Sensitivity of Test C",
       y = "Sensitivity",
       x = "Method") +
  theme_minimal()
```

```{r}

##Testing to see if wald is giving me the right values

# Compute Wald Confidence Interval for Sensitivity
wald_ci_lower <- Sensitivity - 1.96 * sqrt((Sensitivity * (1 - Sensitivity)) / (TP + FN))
wald_ci_upper <- Sensitivity + 1.96 * sqrt((Sensitivity * (1 - Sensitivity)) / (TP + FN))

wald_ci_lower <- max(0, wald_ci_lower)
wald_ci_upper <- min(1, wald_ci_upper)

print(paste("Wald CI for Sensitivity: [", round(wald_ci_lower, 4), ",", round(wald_ci_upper, 4), "]"))
```

All three methods are close to each other.

Wald CI and Wilson CI have very similar widths and centers. Clopper-Pearson CI is slightly wider due to its exact nature, being more conservative, especially with smaller sample sizes or proportions near the boundaries (0 or 1).

------------------------------------------------------------------------

### Exercise 3: Composite reference standard

Repeat Exercise 2 considering for imperfect reference test:

1.  Test A and B combined with the "and" rule.
2.  Test A and B combined with the "or" rule.

Compare and discuss your findings with the previous exercise.

```{r}
# AND Rule (Both Test A and B positive)
TP_and <- sum(data %>% filter(TestA == "Positive" & TestB == "Positive" & TestC == "Positive") %>% pull(Frequency))
FN_and <- sum(data %>% filter(TestA == "Positive" & TestB == "Positive" & TestC == "Negative") %>% pull(Frequency))
total_and <- TP_and + FN_and

# OR Rule (Either Test A or Test B positive)
TP_or <- sum(data %>% filter((TestA == "Positive" | TestB == "Positive") & TestC == "Positive") %>% pull(Frequency))
FN_or <- sum(data %>% filter((TestA == "Positive" | TestB == "Positive") & TestC == "Negative") %>% pull(Frequency))
total_or <- TP_or + FN_or

# Compute Sensitivity for AND and OR rules
sensitivity_and <- TP_and / total_and
sensitivity_or <- TP_or / total_or

# Compute Confidence Intervals
wald_ci_and <- binom.confint(TP_and, total_and, method = "asymptotic")
clopper_pearson_ci_and <- binom.confint(TP_and, total_and, method = "exact")
wilson_ci_and <- binom.confint(TP_and, total_and, method = "wilson")

wald_ci_or <- binom.confint(TP_or, total_or, method = "asymptotic")
clopper_pearson_ci_or <- binom.confint(TP_or, total_or, method = "exact")
wilson_ci_or <- binom.confint(TP_or, total_or, method = "wilson")

# Create DataFrame for visualization
ci_data <- data.frame(
  Method = rep(c("Wald", "Clopper-Pearson", "Wilson"), 2),
  Rule = c(rep("AND", 3), rep("OR", 3)),
  Lower = c(wald_ci_and$lower, clopper_pearson_ci_and$lower, wilson_ci_and$lower, 
            wald_ci_or$lower, clopper_pearson_ci_or$lower, wilson_ci_or$lower),
  Upper = c(wald_ci_and$upper, clopper_pearson_ci_and$upper, wilson_ci_and$upper, 
            wald_ci_or$upper, clopper_pearson_ci_or$upper, wilson_ci_or$upper)
)

# Visualization of confidence intervals
ggplot(ci_data, aes(x = Method, y = (Lower + Upper) / 2, color = Rule)) +
  geom_point(size = 3, position = position_dodge(width = 0.5)) +
  geom_errorbar(aes(ymin = Lower, ymax = Upper), width = 0.2, position = position_dodge(width = 0.5)) +
  labs(title = "95% Confidence Intervals for Sensitivity of Test C (Composite Reference)",
       y = "Sensitivity",
       x = "Method") +
  theme_minimal()

```

When applying the AND rule, the estimated sensitivity of Test C was noticeably higher compared to using Test A alone. This is because the AND rule results in a more specific reference standard, meaning fewer cases qualify as "truly positive". It reduces the number of false positives in the reference, leading to a higher estimated sensitivity for Test C. However, this also means that the reference standard is more likely to miss true positives cases, increasing the number of false negatives.

On the other hand, the OR rule produced a much lower estimated sensitivity for Test C. Since the OR rule includes any case where either Test A or Test B is positive, the total number of reference positives increases. It reduces false negatives in the reference standard but increases false positives, leading to a lower sensitivity estimate for Test C. The confidence intervals for the OR rule were also wider, indicating more uncertainty in the sensitivity estimate.

Compared to using Test A alone, the findings confirm that the choice of reference standard has a significant impact on the estimated sensitivity of Test C. The AND rule results in a higher sensitivity estimate, suggesting that Test C appears to perform better under a stricter reference standard. In contrast, the OR rule results in a lower sensitivity estimate, showing that Test C performs worse when the reference is more inclusive. The sensitivity estimate using only Test A as a reference falls between the AND and OR results, indicating that a single test reference may not be the most reliable approach.

------------------------------------------------------------------------

### Exercise 4: Confidence intervals for the relative risk

Compute the relative risk of Test B and C (using Test A as an imperfect reference) and the associated 99% approximate confidence interval. Is there any statistical evidence that the relative risk is one, at a 1% significance level? Visualize your findings adding a line marking the value of a relative risk equal to 1.

Interpret your findings.

```{r}

library(tidyverse)
library(ggplot2)

# Create a contingency table for Test B and Test C
B_pos_C_pos <- sum(data %>% filter(TestB == "Positive" & TestC == "Positive") %>% pull(Frequency))
B_pos_C_neg <- sum(data %>% filter(TestB == "Positive" & TestC == "Negative") %>% pull(Frequency))
B_neg_C_pos <- sum(data %>% filter(TestB == "Negative" & TestC == "Positive") %>% pull(Frequency))
B_neg_C_neg <- sum(data %>% filter(TestB == "Negative" & TestC == "Negative") %>% pull(Frequency))

# Compute relative risk
risk_B_pos <- B_pos_C_pos / (B_pos_C_pos + B_pos_C_neg)
risk_B_neg <- B_neg_C_pos / (B_neg_C_pos + B_neg_C_neg)
relative_risk <- risk_B_pos / risk_B_neg

# Compute 99% confidence interval for RR
ln_rr <- log(relative_risk)
se_ln_rr <- sqrt((1 / B_pos_C_pos) - (1 / (B_pos_C_pos + B_pos_C_neg)) + (1 / B_neg_C_pos) - (1 / (B_neg_C_pos + B_neg_C_neg)))
z_value <- qnorm(1 - 0.01 / 2)  # 99% CI (z-score for 99% confidence)
ci_lower <- exp(ln_rr - z_value * se_ln_rr)
ci_upper <- exp(ln_rr + z_value * se_ln_rr)

# Determine statistical significance
statistical_evidence <- ifelse(ci_lower <= 1 & ci_upper >= 1, "No", "Yes")

# Create a data frame for visualization
rr_data <- data.frame(
  Method = "Relative Risk",
  RR = relative_risk,
  Lower = ci_lower,
  Upper = ci_upper
)

# Visualization
ggplot(rr_data, aes(x = Method, y = RR)) +
  geom_point(size = 4, color = "blue") +
  geom_errorbar(aes(ymin = Lower, ymax = Upper), width = 0.2, color = "black") +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red", size = 1) +
  labs(title = "99% Confidence Interval for Relative Risk (Test B vs. Test C)",
       y = "Relative Risk",
       x = "") +
  theme_minimal()

# Print results
cat("Relative Risk:", round(relative_risk, 4), "\n")
cat("99% Confidence Interval: (", round(ci_lower, 4), ",", round(ci_upper, 4), ")\n")
cat("Statistical Evidence that RR ≠ 1 at 1% level:", statistical_evidence, "\n")

```

The results of this analysis provide strong statistical evidence of an association between Test B and Test C, using Test A as an imperfect reference standard. The estimated relative risk is 15.28, meaning that individuals who test positive in Test B are approximately 15 times more likely to also test positive in Test C compared to those who test negative in Test B.

The 99% confidence interval for the relative risk ranges from 9.25 to 25.25. We can reject the null hypothesis (RR = 1), which states that there is no association between Test B and Test C. The fact that the entire confidence interval lies well above 1 confirms that individuals with a positive result in Test B have a much higher likelihood of testing positive in Test C.
